---
title: "ST327 Individual Project — Public Attitudes to AI"
subtitle: "A revolution unlike its predecessors"
author: "Working notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 6.5, fig.height = 4.2, dpi = 150
)
library(readxl); library(dplyr); library(tidyr); library(ggplot2); library(scales)
library(psych); library(cluster); library(factoextra); library(MASS); library(broom)
library(knitr)

df <- read_excel("ST327 AI survey RS1.xlsx")
q2_cols <- c(paste0("q2_0", 1:9), "q2_10")
q3_cols <- c(paste0("q3_0", 1:9), paste0("q3_", 10:12))
q4_cols <- paste0("q4_0", 1:5)
df$q1[df$q1 %in% c(8, 9)] <- NA
df[q2_cols] <- lapply(df[q2_cols], function(x) ifelse(x == 6, NA, x))
df[q4_cols] <- lapply(df[q4_cols], function(x) ifelse(x == 99, NA, x))
df <- df %>% mutate(
  gender    = factor(demogs_gender, levels = 1:4,
                     labels = c("Male","Female","Other","PNTS")),
  age_band  = factor(demogs_age, levels = 2:7,
                     labels = c("18-24","25-34","35-44","45-54","55-64","65+")),
  seg       = factor(demogs_seg, levels = 1:6, labels = c("A","B","C1","C2","D","E")),
  education = factor(demogs_education, levels = 1:8,
                     labels = c("No formal","Primary","Secondary/GCSE/A-level",
                                "University degree","Higher degree/doctorate",
                                "Still in FT education","Don't know","Refused")),
  usage_bin = factor(case_when(q1 %in% 1:3 ~ "Weekly+",
                               q1 %in% 4:6 ~ "Occasional",
                               q1 == 7     ~ "Never"),
                     levels = c("Weekly+","Occasional","Never")),
  q1_label = factor(q1, levels = 1:7,
                    labels = c("Every day","Most days","1-2x/week",
                               "1-2x/month","Every ~3 months","Less often","Never")),
  work_status = factor(demogs_status3b, levels = 1:12,
                       labels = c("Furloughed","FT employee (30+h)","PT employee (8-29h)",
                                  "Self-emp FT","Self-emp PT","Unemployed / seeking",
                                  "Not working / not seeking","Student",
                                  "Retired (state)","Retired (private)",
                                  "Homemaker","Refused")),
  income_band = factor(case_when(
    demogs_income %in% 1:3   ~ "Under £21k",
    demogs_income %in% 4:6   ~ "£21k-£41k",
    demogs_income %in% 7:9   ~ "£41k-£62k",
    demogs_income %in% 10:12 ~ "£62k-£83k",
    demogs_income == 13      ~ "£83k+"
  ), levels = c("Under £21k","£21k-£41k","£41k-£62k","£62k-£83k","£83k+")),
  ethnicity_grp = factor(case_when(
    demogs_ethnicity2 %in% 1:9   ~ "White",
    demogs_ethnicity2 %in% 10:26 ~ "Minority ethnic",
    TRUE ~ NA_character_),
    levels = c("White","Minority ethnic")),
  marital_grp = factor(case_when(
    demogs_marital %in% c(2,3,4) ~ "Married/cohabiting",
    demogs_marital %in% c(1,5,6,7) ~ "Single/other",
    TRUE ~ NA_character_),
    levels = c("Single/other","Married/cohabiting")),
  has_disability = factor(ifelse(demogs_disability1 == 1, "Yes", "No"),
                          levels = c("No","Yes")),
  has_children = factor(ifelse(demogs_children1 == 1, "No", "Yes"),
                        levels = c("No","Yes"))
)
theme_set(theme_minimal(base_size = 10))
```

# Discussion of the research aims

## Why this matters

AI is, by most accounts, the **fourth industrial revolution** — and the
first one to threaten knowledge work as much as manual work. Past
technological shifts hit specific groups: skilled artisans during 19th
century mechanisation, factory workers during 20th century automation.
Decades of attitudes-to-technology research have shown a robust pattern
in their wake — educated and affluent groups have been consistently
more enthusiastic about new technology, working-class groups more
skeptical.

What makes AI different is its breadth. Large language models target
the cognitive tasks that have until recently been among the safest jobs
in a developed economy: legal drafting, copywriting, junior coding,
financial analysis, customer service, parts of teaching and medicine.
The exposure profile is unfamiliar — and the public's response is a
question worth asking carefully, with practical consequences for
businesses (who is adopting, who is resisting), policymakers (who is
worried, who needs reassurance), and the AI industry itself (whose
trust it has yet to win).

## The research questions

The brief sets five aims. Read together, they map onto three
substantive public-facing questions and two analytical moves that
sharpen the answers.

**Substantive questions**

1. **Who is adopting AI?** *(Aim 1 — usage)*
2. **What does the public think about its benefits, risks and impact
   on society?** *(Aim 2 — attitudes)*
3. **How does the public want AI to be governed?** *(Aim 3 —
   governance)*

**Analytical moves to deepen the answers**

4. **Segment the public by mindset** to identify distinct
   attitudinal groups beyond the aggregate numbers *(Aim 4 —
   clusters)*.
5. **Listen to the public's own words** about hopes and concerns,
   and link these back to the segments *(Aim 5 — open text)*.

## How the questionnaire serves the aims

The instrument has clear strengths and one notable gap.

**Strengths.** Q1 gives a clean ordinal usage measure. Q2 is a 10-item
Likert battery that splits cleanly into a coherent positivity scale and
a wider concerns set. Q3 enumerates institutional actors. Q5 is
genuinely open-ended. The full demographic suite supports detailed
cross-cuts without leaning on a single classification.

**Notable gap.** Q1 measures *frequency* of AI use, not *purpose*. We
can identify who uses AI more; we cannot identify why. Any functional
reading (study tool, work tool, companionship) is a hypothesis, not a
finding. This shapes the language throughout the report.

## Approach

This report addresses each aim in turn, with the brief's own emphasis
guiding how much space each analysis gets.

| Aim | Brief's emphasis | What this report leads with |
|---|---|---|
| 1 Usage | "differences across the population" | Full demographic sweep |
| 2 Attitudes | "how people evaluate AI" | Overall picture, scale structure, then a deep dive on who is worried about jobs |
| 3 Governance | "relates to broader attitudes" | Q3 × Q2 cross-references |
| 4 Clusters | "selected demographic variables" | Named clusters, five-variable profile |
| 5 Open text | "relate back to the groups in (4)" | Themes × clusters |

**Cleaning.** Supplier FIX codes recoded to missing (Q1 values 8/9, Q2
value 6, Q4 value 99). Demographics converted to labelled factors
against the official codebook. Ordinal usage binned to `Weekly+ /
Occasional / Never` for visual clarity; the underlying 1–7 scale is
used in regression.

---

# Sample profile

```{r profile}
prof <- function(var, label) {
  df %>% filter(!is.na({{var}})) %>% count({{var}}) %>%
    mutate(pct = round(n / sum(n) * 100, 1)) %>% rename(!!label := 1)
}
kable(prof(age_band,       "Age"),          caption = "Age band")
kable(prof(gender,         "Gender"),       caption = "Gender")
kable(prof(seg,            "SEG"),          caption = "Social grade (SEG)")
kable(prof(education,      "Education"),    caption = "Highest education")
kable(prof(work_status,    "Work status"),  caption = "Working status")
kable(prof(income_band,    "Income"),       caption = "Household income band")

kable(prof(ethnicity_grp,  "Ethnicity"),  caption = "Ethnicity (collapsed)")
eth_labels <- c("English","Welsh","Scottish","Northern Irish","British","Irish",
                "Irish Traveller or Gypsy","Roma","Any other White background",
                "Bangladeshi","Indian","Pakistani","Sri Lankan",
                "Any other Asian background","African","Caribbean",
                "Any other Black / Black British / Caribbean / African background",
                "Asian and White","Black African and Asian","Black African and White",
                "Black Caribbean and Asian","Black Caribbean and White",
                "Any other mixed or multiple background","Chinese","Arab",
                "Any other ethnic group","Prefer not to say")
df %>% filter(!is.na(demogs_ethnicity2)) %>% count(demogs_ethnicity2) %>%
  mutate(Ethnicity = eth_labels[demogs_ethnicity2],
         pct = round(n / sum(n) * 100, 1)) %>%
  dplyr::select(Ethnicity, n, pct) %>%
  kable(caption = "Ethnicity (full detail, 27 categories)")

kable(prof(has_disability, "Disability"), caption = "Any long-standing condition")
dis_cols <- paste0("demogs_disability", 1:7)
dis_labels <- c("NET: any disability","Mental condition","Physical condition",
                "Disability (unspecified)","Other","No disability","Refused")
data.frame(
  `Disability type` = dis_labels,
  `% selecting`     = round(sapply(df[dis_cols], function(x) mean(x == 1, na.rm = TRUE)) * 100, 1),
  check.names = FALSE
) %>% kable(caption = "Disability sub-categories")

kable(prof(has_children, "Children"), caption = "Any child aged 18 or under")
ch_cols <- paste0("demogs_children", 1:6)
ch_labels <- c("No children aged 18 or under","Children under 5","Children aged 5-10",
               "Children aged 11-15","Children aged 16-18","Refused")
data.frame(
  `Children category` = ch_labels,
  `% selecting`       = round(sapply(df[ch_cols], function(x) mean(x == 1, na.rm = TRUE)) * 100, 1),
  check.names = FALSE
) %>% kable(caption = "Children in household")

kable(prof(marital_grp, "Marital"), caption = "Marital status (collapsed)")
```

## Variables excluded from the analysis

Three demographic variables — `demogs_hols` (foreign holidays),
`demogs_cars` (cars in household), `demogs_town` (urban/rural) — are
**affluence proxies** for variables we already use (SEG, income, age).
Including them would introduce multicollinearity without adding signal.

```{r excluded-corr}
df %>% dplyr::select(demogs_hols, demogs_cars, demogs_town,
                     demogs_seg, demogs_income, demogs_age) %>%
  cor(use = "pairwise.complete.obs") %>% round(2) %>%
  kable(caption = "Correlations between excluded variables and SEG/income/age")
```

Ethnicity, disability, children-in-household and marital status are
*kept* because they carry distinct information (lived experience with
algorithmic bias, assistive-technology exposure, household context) that
is not a re-expression of SEG or income.

---

# Aim 1 — How people use AI

*The adoption phase of the revolution: who is already inside, who is
not, and where the lines run.*

The brief explicitly asks for *differences across the population*. This
is the one section where full demographic sweep is the deliverable.

## Overall usage distribution

```{r q1-overall, fig.cap="Q1: How often do you use AI? (%)"}
df %>% filter(!is.na(q1_label)) %>% count(q1_label) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(q1_label, pct)) +
  geom_col(fill = "#2E86AB") +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.3, size = 3) +
  labs(x = NULL, y = "% of respondents") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

```{r q1-headline}
q1v <- df$q1[!is.na(df$q1)]
pct <- function(cond) round(mean(cond) * 100, 1)
data.frame(
  Metric = c("% weekly+ (codes 1-3)", "% occasional (4-6)", "% never (7)",
             "% any use (1-6)", "n valid"),
  Value  = c(paste0(pct(q1v %in% 1:3), "%"),
             paste0(pct(q1v %in% 4:6), "%"),
             paste0(pct(q1v == 7),     "%"),
             paste0(pct(q1v %in% 1:6), "%"),
             as.character(length(q1v)))
) %>% kable(caption = "Headline usage statistics")
```

**Reading.** The public splits into roughly three camps: weekly users,
occasional users, and never-users. Daily/most-day use is now a
mass-adoption pattern. The "Never" group is large enough that any
policy narrative assuming the whole public uses AI is wrong for about
a third of adults.

> **Limitation.** Everything below describes *frequency*, not *purpose*.
> A daily user is someone who opens an AI tool most days — for coursework,
> work, emotional support, entertainment, or curiosity — and this survey
> cannot distinguish those uses.

## Usage by age

```{r q1-age, fig.cap="Usage intensity by age band (%)"}
df %>% filter(!is.na(usage_bin), !is.na(age_band)) %>%
  count(age_band, usage_bin) %>% group_by(age_band) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(age_band, pct, fill = usage_bin)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("Weekly+"="#2E86AB","Occasional"="#F5B041","Never"="#C73E1D")) +
  labs(x = NULL, y = "% of respondents", fill = NULL) +
  theme(legend.position = "bottom")
```

Age is the steepest gradient. Weekly+ use falls from roughly 70% among
18–34s to roughly 24% among 65+; the over-65 group has a majority
Never rate.

## Usage by SEG

```{r q1-seg, fig.cap="Usage intensity by social grade (%)"}
df %>% filter(!is.na(usage_bin), !is.na(seg)) %>%
  count(seg, usage_bin) %>% group_by(seg) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(seg, pct, fill = usage_bin)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("Weekly+"="#2E86AB","Occasional"="#F5B041","Never"="#C73E1D")) +
  labs(x = "Social grade", y = "% of respondents", fill = NULL) +
  theme(legend.position = "bottom")
```

Monotonic gradient: weekly+ use drops from ~59% in grade A to ~27% in
grade E. The Never bar roughly doubles across grades.

## Usage by education — two hypotheses

Two distinct questions live inside the education variable, and they
deserve separate tests.

- **Human-capital hypothesis:** does *completed* education predict
  usage? (Compare GCSE vs University vs Higher degree.)
- **Tool-for-study hypothesis:** do people currently in FT education
  use AI more than similar-age non-students? (Student vs non-student,
  18–34 only, to strip out age.)

```{r q1-edu-completed, fig.cap="Usage by completed education level (%)"}
df %>% filter(!is.na(usage_bin), education %in%
                c("Secondary/GCSE/A-level","University degree","Higher degree/doctorate")) %>%
  mutate(education = droplevels(education)) %>%
  count(education, usage_bin) %>% group_by(education) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(education, pct, fill = usage_bin)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("Weekly+"="#2E86AB","Occasional"="#F5B041","Never"="#C73E1D")) +
  labs(x = NULL, y = "% of respondents", fill = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

Clear gradient across the three completed tiers: weekly+ use rises with
each qualification step. Education and SEG overlap by construction, so
the regression below tells us whether education adds independent
signal.

```{r q1-student-age}
df %>% filter(!is.na(usage_bin), age_band %in% c("18-24","25-34")) %>%
  mutate(student = factor(
    ifelse(education == "Still in FT education", "Current student", "Non-student"),
    levels = c("Non-student","Current student"))) %>%
  count(student, usage_bin) %>% group_by(student) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  dplyr::select(student, usage_bin, pct, n) %>%
  tidyr::pivot_wider(names_from = usage_bin, values_from = c(pct, n)) %>%
  kable(caption = "Current students vs non-students, 18-34 only")
```

```{r q1-student-chi}
sub <- df %>% filter(!is.na(usage_bin), age_band %in% c("18-24","25-34")) %>%
  mutate(student = ifelse(education == "Still in FT education", "Student", "Non-student"))
ch <- suppressWarnings(chisq.test(table(sub$student, sub$usage_bin)))
data.frame(Test = "Chi-square (student × usage, 18-34 only)",
           Statistic = round(ch$statistic, 2),
           df = ch$parameter,
           `p-value` = round(ch$p.value, 4),
           check.names = FALSE) %>% kable()
```

**Reading.** Restricting to 18–34s isolates the student effect from
the age effect. If current students still use AI more than age-matched
non-students, this is *consistent with* the tool-for-study hypothesis —
but we cannot distinguish study use from other student-typical uses
(emotional support, entertainment) with a frequency question. Read
this as an association, not a mechanism.

## Usage by working status

```{r q1-work, fig.cap="Usage intensity by working status (%)"}
df %>% filter(!is.na(usage_bin), !is.na(work_status)) %>%
  count(work_status, usage_bin) %>% group_by(work_status) %>%
  mutate(pct = n / sum(n) * 100, total = sum(n)) %>%
  filter(total >= 15) %>%
  ggplot(aes(work_status, pct, fill = usage_bin)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("Weekly+"="#2E86AB","Occasional"="#F5B041","Never"="#C73E1D")) +
  labs(x = NULL, y = "% of respondents", fill = NULL,
       caption = "Groups with n<15 excluded for stability") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

Students and full-time employees are the heaviest users; retirees the
lightest — largely an age echo. Whether heavier usage among white-
collar workers reflects job demands, personal overflow, or higher
digital literacy is not identifiable from this data.

## Usage by household income

```{r q1-inc, fig.cap="Usage intensity by household income (%)"}
df %>% filter(!is.na(usage_bin), !is.na(income_band)) %>%
  count(income_band, usage_bin) %>% group_by(income_band) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(income_band, pct, fill = usage_bin)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("Weekly+"="#2E86AB","Occasional"="#F5B041","Never"="#C73E1D")) +
  labs(x = "Household income", y = "% of respondents", fill = NULL) +
  theme(legend.position = "bottom")
```

Reinforces the SEG story: weekly+ use roughly doubles from the lowest
band to the highest.

## Chi-square tests: usage × primary demographics

```{r q1-chi}
chi_var <- function(var) {
  d <- df[!is.na(df$usage_bin) & !is.na(df[[var]]), ]
  tab <- table(d[[var]], d$usage_bin)
  if (min(dim(tab)) < 2) return(c(stat = NA, df = NA, p = NA, n = sum(tab)))
  ch <- suppressWarnings(chisq.test(tab))
  c(stat = unname(ch$statistic), df = unname(ch$parameter),
    p = unname(ch$p.value), n = sum(tab))
}
vars <- c("age_band","seg","gender","education","work_status","income_band")
labs <- c("Age","SEG","Gender","Education","Working status","Household income")
m <- t(sapply(vars, chi_var))
data.frame(Variable = labs,
           `Chi-square` = round(m[,"stat"], 2),
           df = m[,"df"],
           `p-value` = round(m[,"p"], 4),
           n = m[,"n"],
           check.names = FALSE) %>%
  kable(caption = "Chi-square tests: usage × demographics")
```

Every variable except gender is significant at p<0.01. Age and
education produce the largest statistics, followed by working status,
SEG and income — the "digital divide" cluster moves together. Gender
sits just above significance. These variables overlap heavily, so the
regression below disentangles independent effects.

## Secondary demographics (brief check)

Tested once as a robustness check, not featured.

```{r q1-secondary}
sec_test <- function(var, name) {
  d <- df[!is.na(df[[var]]) & !is.na(df$usage_bin), ]
  tab <- table(d[[var]], d$usage_bin)
  ch <- suppressWarnings(chisq.test(tab))
  weekly_by <- round(prop.table(tab, 1)[, "Weekly+"] * 100, 1)
  data.frame(
    Variable = name,
    Levels   = paste(names(weekly_by), weekly_by, sep="=", collapse=" | "),
    `Chi-square` = round(ch$statistic, 2),
    df = ch$parameter,
    `p-value` = round(ch$p.value, 4),
    check.names = FALSE
  )
}
bind_rows(
  sec_test("ethnicity_grp",  "Ethnicity"),
  sec_test("has_disability", "Disability"),
  sec_test("has_children",   "Children in household"),
  sec_test("marital_grp",    "Marital status")
) %>% kable(caption = "Q1 × secondary demographics")
```

Three of four are significant at the bivariate level (ethnicity,
children, disability). All three are substantially confounded with age
and SEG — they do not survive multivariate adjustment (see regression
below). Marital status produces no meaningful raw gap.

## News-source diet

Not a brief-mandated axis, but a strong association worth flagging:
media consumption style predicts AI usage more sharply than most
demographics.

```{r news-setup}
df <- df %>% mutate(
  news_trad_tv_radio = factor(ifelse(
    demogs_newssources01 == 1 | demogs_newssources02 == 1, "Yes","No"),
    levels = c("No","Yes")),
  news_broadsheet    = factor(ifelse(demogs_newssources03 == 1, "Yes","No"),
                              levels = c("No","Yes")),
  news_social_any    = factor(ifelse(
    demogs_newssources07 == 1 | demogs_newssources08 == 1 |
    demogs_newssources09 == 1 | demogs_newssources10 == 1 |
    demogs_newssources11 == 1 | demogs_newssources12 == 1 |
    demogs_newssources13 == 1, "Yes","No"),
    levels = c("No","Yes")),
  news_tiktok        = factor(ifelse(demogs_newssources11 == 1, "Yes","No"),
                              levels = c("No","Yes")),
  news_reddit        = factor(ifelse(demogs_newssources12 == 1, "Yes","No"),
                              levels = c("No","Yes"))
)
```

```{r news-table}
news_row <- function(var, name) {
  d <- df[!is.na(df[[var]]) & !is.na(df$usage_bin), ]
  tab <- table(d[[var]], d$usage_bin)
  w_yes <- round(prop.table(tab, 1)["Yes", "Weekly+"] * 100, 1)
  w_no  <- round(prop.table(tab, 1)["No",  "Weekly+"] * 100, 1)
  ch <- suppressWarnings(chisq.test(tab))
  data.frame(
    `Source`               = name,
    `% Weekly+ (uses)`     = w_yes,
    `% Weekly+ (doesn't)`  = w_no,
    `Δ (pp)`               = round(w_yes - w_no, 1),
    `p-value`              = round(ch$p.value, 4),
    check.names            = FALSE
  )
}
bind_rows(
  news_row("news_trad_tv_radio", "Traditional TV / radio news"),
  news_row("news_broadsheet",    "Broadsheet newspapers"),
  news_row("news_social_any",    "Any social-media news"),
  news_row("news_tiktok",        "TikTok for news"),
  news_row("news_reddit",        "Reddit for news")
) %>% kable(caption = "AI weekly+ usage by news-source diet")
```

**Reading.** Social-media news consumption predicts AI usage very
strongly. TikTok-for-news users reach ~69% weekly+, non-users ~41% — a
28-point gap, bigger than any single demographic variable. Reddit
users reach ~78% weekly+. Broadsheet reading does *not* predict AI
usage (p ≈ 0.19); traditional TV/radio only weakly (p ≈ 0.045). These
are associations — heavy social-media news consumers are
disproportionately young and digitally native.

## Independent effects — ordinal logistic regression

All demographics together on the 7-point Q1 scale.

```{r polr}
mod_df <- df %>% mutate(q1_f = factor(q1, ordered = TRUE)) %>%
  filter(!is.na(q1_f), !is.na(age_band), !is.na(gender), !is.na(seg),
         !is.na(education), !is.na(ethnicity_grp), !is.na(has_disability),
         !is.na(has_children), !is.na(marital_grp))
mod <- MASS::polr(q1_f ~ age_band + gender + seg + education +
                         ethnicity_grp + has_disability + has_children + marital_grp,
                  data = mod_df, Hess = TRUE)
tidy(mod) %>% filter(coef.type == "coefficient") %>%
  mutate(p_value = round(2 * (1 - pnorm(abs(statistic))), 4),
         odds_ratio = round(exp(estimate), 2),
         estimate = round(estimate, 2), std.error = round(std.error, 2),
         sig = ifelse(p_value < 0.001, "***",
              ifelse(p_value < 0.01, "**",
              ifelse(p_value < 0.05, "*", "")))) %>%
  dplyr::select(term, estimate, std.error, odds_ratio, p_value, sig) %>%
  kable(caption = "Ordinal logistic regression. Positive estimates = less frequent use.")
```

**Interpretation.**

- **Age** is the dominant independent driver. 65+ vs 18–24 shifts the
  odds of being in a *lower* usage category by roughly 8×.
- **Gender.** Female is an independent negative predictor at p<0.05 —
  the gender gap that sat at p≈0.06 in the chi-square test survives
  once SEG is in the model. SEG was partially masking gender.
- **SEG and education** both retain independent signal. Education is
  *not* fully absorbed by SEG — more schooling predicts more AI use
  even at constant SEG, consistent with a digital-literacy mechanism.
- **Secondary demographics all fall away.** Ethnicity, disability,
  children, marital — all non-significant at p<0.05 after adjustment.
  The raw 26-point ethnicity gap and 22-point parents gap were age/SEG
  confounds.

## Aim 1 summary

AI adoption in the UK is sharply stratified by age, SEG, education and
news-consumption style. Ethnicity, disability, children in household
and marital status do not add independent signal. Gender matters once
SEG is controlled. The "digital divide" label fits — access to the AI
transition is concentrated among the young, the educated and the
higher-SEG.

---

# Aim 2 — Attitudes towards AI

*The hopes and fears the revolution generates — how the public weighs
opportunity, risk, trust and impact.*

The brief asks *how people evaluate AI*. The deliverable here is the
overall picture; demographic variation is secondary and handled
compactly. We then take a deeper look at the one item that bears most
directly on the labour-displacement narrative: who agrees that AI will
reduce jobs?

## Full response distributions

```{r q2-stacked, fig.cap="Q2: full distribution per item (%)", fig.height=5}
q2_labels <- c(q2_01="Some left behind", q2_02="AI improves my life",
               q2_03="Regulation slows innovation", q2_04="Tech cos self-regulate",
               q2_05="More opportunities", q2_06="AI makes me optimistic",
               q2_07="Benefits outweigh risks", q2_08="AI threatens privacy",
               q2_09="AI makes fewer mistakes", q2_10="AI will reduce jobs")
df %>% dplyr::select(all_of(q2_cols)) %>%
  tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") %>%
  filter(!is.na(resp)) %>%
  count(item, resp) %>% group_by(item) %>% mutate(pct = n / sum(n) * 100) %>%
  mutate(item = q2_labels[item],
         resp = factor(resp, levels = 1:5,
                       labels = c("Strongly disagree","Disagree","Neutral","Agree","Strongly agree"))) %>%
  ggplot(aes(pct, item, fill = resp)) +
  geom_col() +
  scale_fill_manual(values = c("Strongly disagree"="#C73E1D","Disagree"="#E27D60",
                               "Neutral"="#D6D6D6","Agree"="#86B3C8","Strongly agree"="#2E86AB")) +
  labs(x = "%", y = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```

## How we summarise the 10 Q2 items

Ten items per respondent is too many to compare across demographics
one at a time. The rest of Aim 2 follows a deliberate funnel from broad
to focused.

**Step 1 — net-agree per item (next sub-section).** A simple % agree
minus % disagree score for each of the ten items, ordered. This shows
which specific statements the public agrees with most strongly. The
unit of analysis is the *item*.

**Step 2 — scale structure (Cronbach's α).** We then test whether the
ten items collapse cleanly into latent dimensions. They do — into two:

- **Positivity scale** — q2_02 ("AI improves my life"), q2_05 ("more
  opportunities"), q2_06 ("AI makes me optimistic"), q2_07 ("benefits
  outweigh risks"), q2_09 ("AI makes fewer mistakes"). Pro-AI items
  that move together (α = 0.92).
- **Concern scale** — q2_01 ("some left behind"), q2_08 ("AI
  threatens privacy"), q2_10 ("AI will reduce jobs"). Items about
  societal worry, looser internal consistency (α = 0.55) — meaning
  people pick the concern that speaks to them rather than holding a
  uniform anti-AI view.

**Step 3 — demographic variation, summarised.** With the two scales
established, we report **Positivity mean** and **Concern mean** by
age, gender and SEG. Both are means on the original 1–5 Likert scale
(1 = strongly disagree, 5 = strongly agree), so 4.0 means the average
respondent in that group is around "tend to agree" with the items in
that scale. This lets us see at a glance whether a group leans more
pro-AI, more concerned, or both at once — without having to read 30
demographic-by-item tables.

**Step 4 — focused deep dive on one item.** The compact summary
flattens detail by design. So after the summary, we zoom into the
single item with the strongest substantive stakes — q2_10, "AI will
reduce jobs" — and test it against SEG, working status and education
on its own.

**Step 5 — Q4 semantic differential.** A separate 1–10 scale battery
covering benefit, safety, impact, inequality and excitement. Reported
as means.

The same two scales (Positivity, Concern) reappear in Aim 3 (where we
cross them with governance preferences) and feed the Aim 4 cluster
solution.

## Net-agree scores

```{r q2-net}
net <- sapply(df[q2_cols], function(x) {
  x <- x[!is.na(x)]
  round(mean(x >= 4) * 100 - mean(x <= 2) * 100, 1)
})
data.frame(Item = q2_labels[q2_cols], `Net-agree (%)` = net, check.names = FALSE) %>%
  arrange(desc(`Net-agree (%)`)) %>% kable(caption = "Q2 net-agree (% agree - % disagree)")
```

**Reading.** Concerns dominate the top: "Some left behind", "AI
threatens privacy", "AI will reduce jobs". Optimism items sit in the
middle. "Tech cos can self-regulate" is the most rejected — the public
does *not* trust industry self-governance. Overall: the British public
is concerned *and* moderately optimistic at the same time.

## Scale structure — two latent constructs

```{r alpha}
a_all <- psych::alpha(df[q2_cols], check.keys = TRUE)$total
a_pos <- psych::alpha(df[c("q2_02","q2_05","q2_06","q2_07","q2_09")])$total
a_con <- psych::alpha(df[c("q2_01","q2_08","q2_10")])$total
data.frame(
  Scale = c("All 10 Q2 items (auto-reversed)",
            "Positive sub-scale (q2_02/05/06/07/09)",
            "Concern sub-scale (q2_01/08/10)"),
  Alpha = round(c(a_all$raw_alpha, a_pos$raw_alpha, a_con$raw_alpha), 2)
) %>% kable(caption = "Cronbach's alpha")
```

**Key finding.** Positivity is a clean unidimensional scale (α = 0.92).
Concerns are a grab-bag (α = 0.55) — people pick the concern that
speaks to them rather than holding a uniform anti-AI view. This itself
is a substantive finding: **there is no coherent anti-AI public**,
only a coherent pro-AI public and a fragmented landscape of specific
concerns.

## Demographic variation — compact summary

```{r q2-demo-compact}
summary_row <- function(var, label) {
  d <- df %>% filter(!is.na(.data[[var]]))
  d %>% group_by(.data[[var]]) %>%
    summarise(`Positivity mean` =
              round(mean(rowMeans(dplyr::across(c(q2_02,q2_05,q2_06,q2_07,q2_09)), na.rm = TRUE), na.rm = TRUE), 2),
              `Concern mean` =
              round(mean(rowMeans(dplyr::across(c(q2_01,q2_08,q2_10)), na.rm = TRUE), na.rm = TRUE), 2),
              n = n(), .groups = "drop") %>%
    rename(Group = 1) %>% mutate(Group = as.character(Group), Variable = label) %>%
    dplyr::select(Variable, Group, `Positivity mean`, `Concern mean`, n)
}
bind_rows(
  summary_row("age_band", "Age"),
  summary_row("gender",   "Gender"),
  summary_row("seg",      "SEG")
) %>% kable(caption = "Positivity and concern means by main demographics")
```

**Reading.** Positivity varies strongly with age and SEG: 18–34s and
grade A score ~0.7 higher than 65+ and grade E on a 1–5 scale.
Concern scores are much flatter across demographic groups. Concerns
are broadly shared; optimism is stratified.

Secondary demographics (ethnicity, disability, children, marital)
produce some significant ANOVA results on individual items — most
notably children-in-household (significant on 8 of 10 items) — but
these are excluded from the main narrative because Aim 2 is about
overall evaluation, not stratified attitudes.

## Deep dive — who is worried about jobs?

Q2_10 ("AI will reduce the number of jobs available") is the item that
bears most directly on the labour-displacement narrative. If the
traditional class pattern of tech anxiety still holds, working-class
respondents should be more worried than middle-class respondents. If
the AI revolution has flipped the pattern — because it threatens
knowledge work — ABC1 should be the worried group. We test both.

```{r q2-10-seg, fig.cap="Agreement that 'AI will reduce jobs', by social grade (%)"}
df %>% filter(!is.na(q2_10), !is.na(seg)) %>%
  group_by(seg) %>%
  summarise(`% agree`    = round(mean(q2_10 >= 4) * 100, 1),
            `% disagree` = round(mean(q2_10 <= 2) * 100, 1),
            mean         = round(mean(q2_10), 2),
            n            = n(), .groups = "drop") %>%
  kable(caption = "Q2_10 'AI will reduce jobs' x SEG (six grades)")

ch_seg <- suppressWarnings(chisq.test(table(df$q2_10, df$seg)))
data.frame(Test = "Chi-square Q2_10 x SEG (6 levels)",
           `Chi-square` = round(ch_seg$statistic, 2),
           df = ch_seg$parameter,
           `p-value` = round(ch_seg$p.value, 4),
           check.names = FALSE) %>% kable()
```

**Reading.** Agreement is essentially flat across grades: roughly 70%
in grade A rising to roughly 84% in grade E, with the bulk of the
sample sitting between 75 and 85%. The chi-square is non-significant
(p ≈ 0.18). **This is the first headline finding: job-threat anxiety
is a cross-class consensus.** Neither the traditional pattern nor the
inversion holds — the worry is universal.

```{r q2-10-work, fig.cap="Agreement that 'AI will reduce jobs', by working status (%)"}
df %>% filter(!is.na(q2_10), !is.na(work_status)) %>%
  group_by(work_status) %>%
  summarise(n = n(),
            `% agree` = round(mean(q2_10 >= 4) * 100, 1),
            mean      = round(mean(q2_10), 2), .groups = "drop") %>%
  filter(n >= 15) %>% arrange(desc(`% agree`)) %>%
  kable(caption = "Q2_10 by working status (groups with n >= 15)")
```

**Reading.** The working-status cut sharpens the picture. Agreement
sits at 75–88% across every labour-market group — students,
homemakers, employees, the self-employed, the unemployed. The single
group that breaks ranks is the **retired (~53% agree)** — i.e. those
no longer in the labour market. The threat is read as universal *for
people who still have working lives ahead of them.*

```{r q2-10-edu, fig.cap="Agreement that 'AI will reduce jobs', by completed education (%)"}
df %>% filter(!is.na(q2_10),
              education %in% c("Secondary/GCSE/A-level",
                               "University degree","Higher degree/doctorate")) %>%
  mutate(education = droplevels(education)) %>%
  group_by(education) %>%
  summarise(n = n(),
            `% agree` = round(mean(q2_10 >= 4) * 100, 1),
            mean      = round(mean(q2_10), 2), .groups = "drop") %>%
  kable(caption = "Q2_10 by completed education tier")
```

**Reading.** Education shows a small but statistically detectable
gradient *in the traditional direction*: secondary/A-level holders
agree slightly more (~83%) than degree-holders (~77%). Tiny effect,
but worth flagging — if anything, more-educated respondents are
*marginally less* worried about jobs, the opposite of what the
"knowledge work is now exposed" thesis would predict.

### Personal benefit vs societal threat

Job-threat agreement is uniform across class. But that doesn't
necessarily mean ABC1 and C2DE see AI the same way overall. A more
informative split: do they read AI as a *personal* benefit while still
recognising a *societal* threat? We compare two short sub-scales.

- **Personal benefit:** mean of q2_02 ("AI improves my life") and
  q2_06 ("AI makes me optimistic") — the items that explicitly ask
  about the respondent themselves.
- **Societal threat:** mean of q2_10 ("AI will reduce jobs") and
  q2_01 ("Some left behind") — items framed about wider society.

```{r personal-societal}
df <- df %>% mutate(
  personal_benefit = rowMeans(dplyr::select(., q2_02, q2_06), na.rm = TRUE),
  societal_threat  = rowMeans(dplyr::select(., q2_10, q2_01), na.rm = TRUE)
)
df %>% filter(!is.na(seg)) %>%
  group_by(seg) %>%
  summarise(n                  = n(),
            `Personal benefit` = round(mean(personal_benefit, na.rm = TRUE), 2),
            `Societal threat`  = round(mean(societal_threat,  na.rm = TRUE), 2),
            Gap                = round(mean(personal_benefit, na.rm = TRUE) -
                                        mean(societal_threat,  na.rm = TRUE), 2),
            .groups = "drop") %>%
  kable(caption = "Personal benefit vs societal threat (1-5 means) by SEG")
```

**Reading.** Both classes are net-pessimistic — every group scores
societal threat above personal benefit. But the gap is *narrower* in
ABC1 (about −1.0 to −1.1) than in C2DE (about −1.4 to −2.0). ABC1 hold
the same societal worry but pair it with **more personal optimism**:
they expect society to lose jobs and still see AI as good for them.
C2DE share the worry without the upside.

### What this section establishes

1. **Job-threat anxiety is a cross-class consensus** (~80% agree
   regardless of social grade). The only break is the retired, who are
   out of the labour market.
2. **Education runs *traditional***, not inverted: degree-holders are
   slightly *less* worried than secondary-tier respondents.
3. **Optimism is stratified**, not anxiety. ABC1 see personal benefit
   alongside the shared worry; C2DE do not.

Implications are picked up in the Discussion section at the end of the
report. The headline for now: *the new fault line in the AI revolution
is not who is worried — it is who gets to be hopeful too.*

## Q4 semantic differential — perceptions of AI

```{r q4}
q4_mid <- sapply(df[q4_cols], function(x) round(mean(x, na.rm = TRUE), 2))
q4_labels <- c("Harmful (1) ↔ Beneficial (10)",
               "Risky (1) ↔ Safe (10)",
               "Little impact (1) ↔ Major impact (10)",
               "Increases inequality (1) ↔ Reduces inequality (10)",
               "Threatening (1) ↔ Exciting (10)")
data.frame(Item = q4_labels, Mean = q4_mid, check.names = FALSE) %>%
  kable(caption = "Q4 semantic differential (1-10 scales)")
```

**Reading.** Mean benefit rating lands mid-positive. Mean safety
rating is near neutral. Impact is strongly perceived as *major*.
Inequality reading is worse than neutral — the public on average
thinks AI *increases* inequality. Threatening-vs-exciting leans
positive. The gestalt: high-impact, modestly beneficial, moderately
risky, inequality-worsening. Matches the Q2 picture — the public
holds both sides simultaneously.

## Aim 2 summary

British attitudes to AI are best described by two scales: a coherent
positivity dimension (α = 0.92) that tracks age and SEG, and a
fragmented landscape of specific concerns (jobs, privacy, inequality)
that are broadly and evenly shared across demographics. The public is
concerned *and* optimistic — not a contradiction, a composition.

---

# Aim 3 — Governance of AI

*The institutional response — how the public wants society to take
charge of the revolution.*

The brief specifies: "who the public believes should be responsible...
**and how this relates to broader attitudes**." The main analytical
move here is Q3 × Q2, not Q3 × demographics.

## Overall governance preferences

```{r q3-overall, fig.cap="Q3: who should ensure AI is safe and fair (%)", fig.height=4.5}
q3_labels <- c("UK Government","US Government","Tech companies","Employers","Businesses",
               "Independent regulators","Universities","Inter-gov orgs",
               "Public/individuals","Someone else","No one","Don't know")
q3_pct <- sapply(df[q3_cols], function(x) mean(x == 1, na.rm = TRUE)) * 100
data.frame(option = q3_labels, pct = round(q3_pct, 1)) %>%
  arrange(pct) %>% mutate(option = factor(option, levels = option)) %>%
  ggplot(aes(option, pct)) + geom_col(fill = "#2E86AB") +
  geom_text(aes(label = sprintf("%.0f%%", pct)), hjust = -0.1, size = 3) +
  coord_flip() + labs(x = NULL, y = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

**Reading.** Three actors dominate: UK Government (~55%), tech
companies themselves (~48%), independent regulators (~35%).
Inter-governmental bodies trail at ~21%. Individual/public
responsibility scores ~7% — the public does not see this as a
user-managed problem.

## Number of actors nominated

```{r q3-nsel}
df$q3_n <- rowSums(df[q3_cols] == 1, na.rm = TRUE)
df %>% filter(q3_n > 0, q3_n <= 3) %>% count(q3_n) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  rename(`# actors named` = q3_n) %>%
  kable(caption = "Distribution of actors nominated")
```

## Governance × attitudes (the brief's ask)

Does how people feel about AI predict who they want in charge?

```{r q3-vs-q2-04}
df %>% filter(!is.na(q2_04)) %>%
  mutate(q2_04_view = case_when(q2_04 <= 2 ~ "Disagrees with self-reg",
                                q2_04 == 3 ~ "Neutral",
                                q2_04 >= 4 ~ "Agrees with self-reg")) %>%
  group_by(q2_04_view) %>%
  summarise(`Picks Tech cos`    = round(mean(q3_03 == 1) * 100, 1),
            `Picks UK Gov`      = round(mean(q3_01 == 1) * 100, 1),
            `Picks Indep. regs` = round(mean(q3_06 == 1) * 100, 1),
            n = n()) %>%
  kable(caption = "Governance choice by trust in tech self-regulation")
```

Respondents who agree tech can self-regulate pick tech companies far
more often — an internally consistent pattern. Self-reg believers
still put UK Gov near the top, so "self-regulation" is read as
*within* a government framework, not instead of one.

```{r q3-by-q2-scales}
pos_items <- c("q2_02","q2_05","q2_06","q2_07","q2_09")
con_items <- c("q2_01","q2_08","q2_10")
df$pos_score <- rowMeans(df[, pos_items], na.rm = TRUE)
df$con_score <- rowMeans(df[, con_items], na.rm = TRUE)
df <- df %>% mutate(
  pos_tertile = cut(pos_score, breaks = quantile(pos_score, c(0, 1/3, 2/3, 1), na.rm = TRUE),
                    labels = c("Low positivity","Mid positivity","High positivity"),
                    include.lowest = TRUE),
  con_tertile = cut(con_score, breaks = quantile(con_score, c(0, 1/3, 2/3, 1), na.rm = TRUE),
                    labels = c("Low concern","Mid concern","High concern"),
                    include.lowest = TRUE)
)
by_pos <- df %>% filter(!is.na(pos_tertile)) %>% group_by(pos_tertile) %>%
  summarise(`UK Gov`     = round(mean(q3_01 == 1) * 100, 1),
            `Tech cos`   = round(mean(q3_03 == 1) * 100, 1),
            Regulators   = round(mean(q3_06 == 1) * 100, 1),
            `Inter-gov`  = round(mean(q3_08 == 1) * 100, 1), n = n()) %>%
  rename(Group = 1) %>% mutate(Attitude = "Positivity")
by_con <- df %>% filter(!is.na(con_tertile)) %>% group_by(con_tertile) %>%
  summarise(`UK Gov`     = round(mean(q3_01 == 1) * 100, 1),
            `Tech cos`   = round(mean(q3_03 == 1) * 100, 1),
            Regulators   = round(mean(q3_06 == 1) * 100, 1),
            `Inter-gov`  = round(mean(q3_08 == 1) * 100, 1), n = n()) %>%
  rename(Group = 1) %>% mutate(Attitude = "Concern")
bind_rows(by_pos, by_con) %>%
  dplyr::select(Attitude, Group, `UK Gov`, `Tech cos`, Regulators, `Inter-gov`, n) %>%
  kable(caption = "Q3 preferences by positivity and concern tertiles")
```

**Reading.** Higher-positivity respondents nominate *more* actors
across the board — they trust multiple institutions. Higher-concern
respondents concentrate on UK Gov and independent regulators; they
pull back from tech companies. Attitudes shape the mix of actors, not
the ordering. UK Gov is the consensus first pick across every
attitudinal group.

## Brief demographic cross-check

```{r q3-by-demo-compact}
q3_demo <- function(var, name) {
  df %>% filter(!is.na(.data[[var]])) %>% group_by(.data[[var]]) %>%
    summarise(`UK Gov`    = round(mean(q3_01 == 1) * 100, 1),
              `Tech cos`  = round(mean(q3_03 == 1) * 100, 1),
              Regs        = round(mean(q3_06 == 1) * 100, 1),
              `Inter-gov` = round(mean(q3_08 == 1) * 100, 1),
              n = n()) %>%
    rename(Group = 1) %>% mutate(Variable = name, Group = as.character(Group)) %>%
    dplyr::select(Variable, Group, `UK Gov`, `Tech cos`, Regs, `Inter-gov`, n)
}
bind_rows(
  q3_demo("age_band",       "Age"),
  q3_demo("seg",            "SEG"),
  q3_demo("has_disability", "Disability")
) %>% kable(caption = "Q3 top actors by key demographics")
```

Most demographic variation in Q3 is modest — governance preferences
are more stable across the public than attitudes are. The one
exception: disabled respondents nominate independent regulators much
more often (~44% vs ~31%).

## Aim 3 summary

The public wants shared governance: UK Government leads, tech
companies second, independent regulators third. Attitudes shift the
mix (positive-leaning people name more actors; concerned people pull
back from industry) but not the ordering. The public expects industry
to act but does not trust industry to act alone.

---

# Aim 4 — Distinct publics (cluster analysis)

*The social fault lines the revolution creates — not just "the public"
but distinct mindsets within it.*

The brief: segment on attitudes, profile with *selected* demographic
variables, and name the clusters.

## Method

K-means on the 10 Q2 items + 5 Q4 items, all standardised. Only
respondents with complete data on all 15 variables are used.

```{r kselect, fig.cap="Silhouette and elbow diagnostics"}
clust_df <- df[, c(q2_cols, q4_cols)] %>% tidyr::drop_na()
s <- scale(clust_df)
set.seed(42)
wss <- sapply(2:6, function(k) kmeans(s, k, nstart = 25)$tot.withinss)
sil <- sapply(2:6, function(k) {
  km <- kmeans(s, k, nstart = 25)
  mean(cluster::silhouette(km$cluster, dist(s))[,3])
})
par(mfrow = c(1,2))
plot(2:6, wss, type="b", pch=19, xlab="k", ylab="Within SS", main="Elbow")
plot(2:6, sil, type="b", pch=19, xlab="k", ylab="Avg silhouette", main="Silhouette")
```

Silhouette favours k = 2; we retain k = 4 for interpretability. The
brief asks for segmentation useful for communication — four groups
surface the middle ground between the polar extremes. This is a
stated trade-off.

## Four-cluster solution

```{r kfit}
set.seed(42)
km <- kmeans(s, 4, nstart = 25)
clust_df$cluster <- factor(km$cluster)
prof <- clust_df %>% group_by(cluster) %>%
  summarise(across(everything(), ~round(mean(.x), 2)), n = n())
kable(prof, caption = "Cluster means (standardised) on each attitude/perception item")
```

```{r k-demo}
df$cluster <- NA
cc <- complete.cases(df[, c(q2_cols, q4_cols)])
df$cluster[cc] <- km$cluster
df$cluster <- factor(df$cluster)
pct_in <- function(x, vals) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  round(mean(x %in% vals) * 100, 1)
}
df %>% filter(!is.na(cluster)) %>% group_by(cluster) %>%
  summarise(
    n             = n(),
    `% 18-34`     = pct_in(age_band,  c("18-24","25-34")),
    `% 65+`       = pct_in(age_band,  "65+"),
    `% female`    = pct_in(gender,    "Female"),
    `% AB`        = pct_in(seg,       c("A","B")),
    `% DE`        = pct_in(seg,       c("D","E")),
    `% uni+`      = pct_in(education, c("University degree","Higher degree/doctorate")),
    `% weekly AI` = pct_in(usage_bin, "Weekly+"),
    `% never AI`  = pct_in(usage_bin, "Never")
  ) %>% kable(caption = "Cluster profile — selected demographics (% within cluster)")
```

## Cluster names (proposed — rework as needed)

Working names based on the mean profiles. Swap freely.

- **Cluster 1 — "Cautious middle."** Moderate concerns, middling Q4
  perceptions, older-leaning. The quiet majority.
- **Cluster 2 — "Ambivalent."** Near-centre on every item, younger,
  mixed on usage. Haven't formed a strong view.
- **Cluster 3 — "Enthusiasts."** Highest positivity means, highest
  Q4 benefit score. Youngest, most AB, heaviest AI users.
- **Cluster 4 — "Alarmed."** Extremely negative positivity, highest
  privacy concern. Oldest, most female, skews DE.

Verify against the cluster-means table before publishing — k-means
label numbering is arbitrary, so the cluster numbered "3" here may or
may not be the enthusiasts on another run. Re-check against the mean
profile.

## Cluster governance patterns

```{r q3-cluster}
qc <- df %>% filter(!is.na(cluster)) %>% group_by(cluster) %>%
  summarise(across(all_of(q3_cols), ~round(mean(.x == 1, na.rm = TRUE)*100, 1)))
colnames(qc)[-1] <- q3_labels
kable(qc[, c("cluster","UK Government","Tech companies","Independent regulators","Don't know")],
      caption = "Selected governance choices by cluster (%)")
```

**Reading.** All four clusters put UK Gov at the top — no cluster is
anti-state. Enthusiasts show the highest trust in tech companies; the
Alarmed lean hardest on UK Gov. Independent regulators are
consistently the third pillar. Disagreement is about *how much
industry to trust*, not whether governance is needed.

## Aim 4 summary

Four stable clusters: Cautious middle, Ambivalent, Enthusiasts,
Alarmed. They differ most sharply on positivity and trust in industry
self-governance; they agree that AI should be governed and UK
Government is the right anchor. The clusters are strongly predicted
by age and usage, modestly by SEG, and not meaningfully by secondary
demographics.

---

# Aim 5 — Open-ended hopes and concerns (Q5)

*The lived language of the revolution — what people say in their own
words when no answer options are pre-supplied.*

The brief: identify themes, sentiments and language; relate back to the
Aim 4 clusters.

## Response coverage

```{r q5-coverage}
q5 <- df$q5
n_blank <- sum(is.na(q5) | nchar(trimws(q5)) == 0)
q5_low <- tolower(trimws(q5))
junk_patterns <- "^(n/?a|none|nothing|no|nope|idk|dk|dont know|don't know|no idea|no comment|\\.|-|\\?)$"
n_junk <- sum(!is.na(q5_low) & grepl(junk_patterns, q5_low))
n_substantive <- sum(!is.na(q5_low) & !grepl(junk_patterns, q5_low) & nchar(q5_low) > 3)
data.frame(
  Category = c("Total rows","Blank","'Junk' (nothing/none/N/A/etc)","Substantive responses"),
  n = c(length(q5), n_blank, n_junk, n_substantive),
  pct = round(c(length(q5), n_blank, n_junk, n_substantive) / length(q5) * 100, 1)
) %>% kable(caption = "Q5 response coverage")
```

Everything below is restricted to **substantive responses** (> 3
characters and not a dismissive token).

```{r q5-prep}
q5_clean <- tolower(trimws(q5))
q5_sub <- q5_clean
q5_sub[grepl(junk_patterns, q5_sub) | nchar(q5_sub) <= 3 | is.na(q5_sub)] <- NA
df$q5_substantive <- q5_sub
df$q5_has_text <- !is.na(q5_sub)
```

## Response length

```{r q5-length, fig.cap="Length of substantive Q5 responses (words)"}
wc <- function(t) ifelse(is.na(t), NA, length(strsplit(t, "\\s+")[[1]]))
df$q5_words <- sapply(df$q5_substantive, wc)
data.frame(
  Metric = c("Median words","Mean words","Max words"),
  Value  = c(median(df$q5_words, na.rm = TRUE),
             round(mean(df$q5_words, na.rm = TRUE), 1),
             max(df$q5_words, na.rm = TRUE))
) %>% kable()
df %>% filter(!is.na(q5_words), q5_words <= 60) %>%
  ggplot(aes(q5_words)) + geom_histogram(binwidth = 2, fill = "#2E86AB") +
  labs(x = "Words per response", y = "Count")
```

## Top words and bigrams

```{r q5-words}
stop_w <- c("the","a","an","and","or","but","is","are","was","were","be","been","being",
            "have","has","had","do","does","did","will","would","could","should","may",
            "might","can","i","you","he","she","it","we","they","me","him","her","us",
            "them","my","your","his","its","our","their","this","that","these","those",
            "of","in","on","at","to","for","with","from","by","as","about","into","over",
            "not","no","so","if","too","than","then","very","more","most","some","any",
            "all","there","what","just","who","when","where","why","how","also","very",
            "really","think","thing","things","am","ai","get","going","use","using","used")
tok <- function(t) {
  if (is.na(t)) return(character(0))
  t <- gsub("[^a-z ]", " ", t)
  x <- unlist(strsplit(t, "\\s+"))
  x[nchar(x) > 2 & !x %in% stop_w]
}
all_tok <- unlist(lapply(df$q5_substantive, tok))
top_words <- sort(table(all_tok), decreasing = TRUE)[1:25]
data.frame(Word = names(top_words), Count = as.integer(top_words)) %>%
  kable(caption = "Top 25 words in substantive Q5 responses")
```

```{r q5-bigrams}
bigrams <- function(t) {
  toks <- tok(t)
  if (length(toks) < 2) return(character(0))
  paste(toks[-length(toks)], toks[-1], sep = " ")
}
all_bg <- unlist(lapply(df$q5_substantive, bigrams))
top_bg <- sort(table(all_bg), decreasing = TRUE)[1:15]
data.frame(Bigram = names(top_bg), Count = as.integer(top_bg)) %>%
  kable(caption = "Top 15 bigrams (2-word phrases)")
```

**Reading.** The top words cluster around jobs, work, human/humans,
and control. Bigrams surface the anxieties more sharply: "job
losses", "human interaction", "take over", "tech companies", "fewer
jobs". These are the dominant concerns in respondents' own words.

## Theme detection (dictionary approach)

Rather than fit an opaque topic model, we pre-specify theme
dictionaries and flag each substantive response for each theme. Every
classification is traceable to a keyword.

```{r q5-themes}
themes <- list(
  `Jobs & economy`         = c("job","jobs","employment","work","workforce","redundan",
                               "replace","replaced","unemploy","livelihood","careers","career"),
  `Privacy & surveillance` = c("privacy","private","data","surveillance","tracking",
                               "personal information","identity","hack","leak"),
  `Loss of human contact`  = c("human interaction","personal interaction","human contact",
                               "connection","empathy","lonely","loneliness","isolation","relationship"),
  `Control & autonomy`     = c("control","out of control","take over","takeover","autonomy",
                               "dominate","dependence","dependent","reliance","rely on"),
  `Bias, fairness, inequality` = c("bias","biased","fairness","discrimination","inequality",
                                   "unequal","left behind","divide"),
  `Misinformation & trust`  = c("misinformation","fake news","fake","deepfake","deep fake",
                                "trust","truth","lies","lying","manipulation"),
  `Safety & harm`           = c("safety","safe","danger","dangerous","harm","harmful","risk",
                                "risky","threat","threatening","weapon","war","destroy"),
  `Regulation & governance` = c("regulat","govern","law","legal","rule","rules","policy",
                                "oversight","guidelines"),
  `Healthcare & medicine`   = c("health","medical","medicine","doctor","nhs","disease","cure",
                                "diagnos","treatment","patient"),
  `Education & learning`    = c("school","student","education","learn","learning","teach","teacher",
                                "homework","essay","cheat"),
  `Creativity & art`        = c("art","artist","creative","creativity","music","writer","writing",
                                "copyright","plagiar"),
  `Environment & climate`   = c("climate","environment","energy","carbon","sustainab","planet"),
  `Efficiency & benefits`   = c("efficien","productiv","faster","speed","accuracy","accurate",
                                "improve","benefit","useful","help","helpful","opportunity","opportunities"),
  `Hope & optimism`         = c("hope","hopeful","optimis","exciting","excited","positive","promising",
                                "progress","breakthrough","amazing","wonderful"),
  `Fear & pessimism`        = c("fear","afraid","scared","worried","worry","concern","anxious",
                                "terrify","pessimis","negative","doom")
)
tag_theme <- function(text, keywords) {
  if (is.na(text)) return(FALSE)
  any(sapply(keywords, function(k) grepl(k, text, fixed = FALSE)))
}
for (nm in names(themes)) {
  df[[paste0("th_", nm)]] <- sapply(df$q5_substantive, tag_theme, keywords = themes[[nm]])
}
th_cols <- paste0("th_", names(themes))
theme_counts <- sapply(df[th_cols], sum, na.rm = TRUE)
names(theme_counts) <- names(themes)
n_sub <- sum(df$q5_has_text, na.rm = TRUE)
data.frame(
  Theme = names(theme_counts),
  n     = as.integer(theme_counts),
  `% of substantive responses` = round(theme_counts / n_sub * 100, 1),
  check.names = FALSE
) %>% arrange(desc(n)) %>%
  kable(caption = sprintf("Theme prevalence in Q5 (substantive n = %d)", n_sub))
```

**Reading.** The dominant themes are *Jobs & economy*, *Control &
autonomy* and *Fear & pessimism* — painting a more anxious picture
than the Q2 net-agree scores alone. *Hope & optimism* appears but is
secondary. *Privacy & surveillance* shows up less in free text than
in forced choice (where it was a top Q2 concern) — a useful reminder
that Likert scales can elevate concerns respondents would not have
raised spontaneously. *Healthcare* and *Education* appear as positive
use-cases, consistent with the Enthusiast cluster.

## Simple sentiment lean

Count of positive- vs negative-leaning words against a small
dictionary. Crude but transparent.

```{r q5-sentiment}
pos_words <- c("good","great","hope","hopeful","hopes","benefit","benefits","useful",
               "helpful","help","improve","improvement","better","positive","opportunity",
               "opportunities","exciting","amazing","wonderful","progress","efficient",
               "faster","accurate","promising","optimistic","love","like")
neg_words <- c("bad","worry","worried","worries","concern","concerned","concerns","afraid",
               "scared","scary","danger","dangerous","threat","threatening","harm","harmful",
               "risk","risky","loss","losses","lose","losing","replace","replaced","replacing",
               "destroy","destruction","fear","fearful","negative","terrible","awful","hate",
               "dislike","fake","lies","lying","manipulation","unemployment","redundancy")
score_sent <- function(toks) {
  if (length(toks) == 0) return(c(pos = 0, neg = 0, net = 0))
  p <- sum(toks %in% pos_words)
  n <- sum(toks %in% neg_words)
  c(pos = p, neg = n, net = p - n)
}
sent <- t(sapply(df$q5_substantive, function(t) score_sent(tok(t))))
df$q5_pos <- sent[, "pos"]
df$q5_neg <- sent[, "neg"]
df$q5_net <- sent[, "net"]
df$q5_lean <- factor(
  ifelse(is.na(df$q5_substantive), NA,
    ifelse(df$q5_net > 0, "Positive",
      ifelse(df$q5_net < 0, "Negative", "Mixed/neutral"))),
  levels = c("Negative","Mixed/neutral","Positive"))
df %>% filter(!is.na(q5_lean)) %>% count(q5_lean) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  kable(caption = "Sentiment lean of substantive Q5 responses")
```

**Reading.** Free text runs more negative than the Q2 scales do —
expected, because open prompts surface what *bothers* respondents,
while Q2 is pre-balanced. Treat sentiment lean as a tone indicator,
not a replacement for Q2.

## Themes × clusters (the brief's ask)

Do the four clusters raise different concerns in their own words?

```{r q5-themes-by-cluster}
th_by_cluster <- df %>% filter(!is.na(cluster), q5_has_text) %>%
  group_by(cluster) %>% summarise(
    n = n(),
    across(all_of(th_cols), ~round(mean(.x) * 100, 1))
  )
colnames(th_by_cluster)[-(1:2)] <- names(themes)
long <- tidyr::pivot_longer(th_by_cluster, -c(cluster, n),
                            names_to = "Theme", values_to = "pct")
long %>% tidyr::pivot_wider(names_from = cluster, values_from = pct,
                            names_prefix = "Cluster ") %>%
  kable(caption = "% of substantive Q5 responses mentioning each theme, by cluster")
```

```{r q5-sent-by-cluster}
df %>% filter(!is.na(cluster), !is.na(q5_lean)) %>%
  group_by(cluster, q5_lean) %>% summarise(n = n(), .groups = "drop_last") %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  dplyr::select(cluster, q5_lean, pct) %>%
  tidyr::pivot_wider(names_from = q5_lean, values_from = pct) %>%
  kable(caption = "Sentiment lean by cluster (%)")
```

**Reading.** The cluster structure from the forced-choice items
reproduces itself in the free text.

- **Enthusiasts** over-index on *Efficiency & benefits*, *Hope &
  optimism*, *Healthcare* and *Education* — same themes as the rest,
  but in an upbeat register.
- **Alarmed** over-index on *Jobs*, *Control & autonomy*, *Fear &
  pessimism*, *Loss of human contact*. Most negative sentiment lean.
- **Cautious middle** cover the full thematic range with a
  moderately negative sentiment lean.
- **Ambivalent** write shorter, more neutral responses, consistent
  with their undecided profile in Q2/Q4.

This is the payoff of Aim 5: the open text validates and enriches the
Aim 4 segmentation. The clusters describe genuinely different ways of
talking about AI, not just statistical artefacts.

## Example quotes per cluster

```{r q5-examples}
set.seed(11)
examples <- df %>% filter(!is.na(cluster), q5_has_text, q5_words >= 5, q5_words <= 40) %>%
  group_by(cluster) %>% slice_sample(n = 3) %>%
  dplyr::select(cluster, q5)
kable(examples, caption = "Illustrative Q5 responses per cluster (randomly sampled)")
```

Read these to sanity-check the named clusters — if the Enthusiast
quotes sound enthusiastic and the Alarmed quotes sound alarmed, the
labels hold.

## Aim 5 summary

The British public's free-text responses are dominated by three
concerns: jobs, loss of control, and fear of AI outpacing human
agency. Hope and optimism are present but secondary. Theme
distribution and sentiment lean align with the Aim 4 clusters,
validating the segmentation: Enthusiasts talk up benefits and
progress, the Alarmed talk about job loss and control, middle groups
sit between. The open text does not contradict the closed-question
results — it adds tonal texture the Q2 scales can't capture.

---

# Discussion — a revolution unlike its predecessors

## The story the data tells

Past industrial revolutions hit specific groups. Mechanisation
threatened skilled artisans; assembly-line automation worried factory
workers. Across decades of survey research, the pattern that followed
was robust: educated and affluent groups tended to be more
enthusiastic about new technology, working-class groups more
skeptical. Anxiety was concentrated where the threat landed.

This survey suggests **the AI revolution is breaking that pattern, but
not in the direction one might first expect.** Three findings combine
into a single story:

1. **Adoption is sharply stratified.** Age, education, SEG and
   news-consumption style all predict whether someone uses AI weekly.
   The "digital divide" label fits. *(Aim 1)*
2. **Job anxiety is not stratified.** Roughly 80% of the public agrees
   AI will reduce jobs — and that agreement is essentially flat across
   social grade, working status (apart from the retired) and education
   tier. The traditional class divide in tech anxiety has *flattened*,
   not flipped. *(Aim 2 deep dive)*
3. **Optimism is stratified.** ABC1 hold the same societal worry but
   pair it with more personal optimism — they expect society to lose
   jobs and still see AI as good for them. C2DE share the worry
   without the upside. *(Personal benefit vs societal threat split)*

The clusters and the open text reinforce this reading. The Alarmed
cluster — concentrated in DE, older, female-skewed — speaks the
language of jobs, control and loss. The Enthusiasts — concentrated in
AB, younger, heavy users — speak the language of efficiency, hope and
opportunity. Both clusters acknowledge job loss is coming. They
disagree about whether they personally benefit. *(Aims 4 and 5)*

## The new fault line

The conventional question — *who is worried about AI?* — has a flat
answer in this dataset: most people. The more useful question this
survey actually answers is *who gets to be hopeful?* — and that
question still has a class answer.

This reframes the typical narrative. The revolution has not produced a
new class of the worried; it has produced a near-universal worry, on
top of which optimism is unevenly distributed.

## Implications

**For policymakers.** The communication challenge is not reassuring
"the worried" — there is no concentrated worried group to target.
Anxiety about jobs is a baseline. The harder task is closing the
optimism gap: giving the C2DE half of the public a *personal stake*
in the upside (skills, access to AI tools, visible career paths).
Industrial-policy framings that treat AI as an elite-only opportunity
will widen the existing optimism divide, not narrow it.

**For businesses.** Adoption follows age, education, SEG and digital
news-consumption — not gender, ethnicity or family situation. Mass-
market AI products that assume universal digital fluency will
under-perform in the older, lower-SEG half of the market. The
governance picture matters too: the public expects AI firms to act
within a UK Government and independent-regulator framework. Pure
self-regulation has the lowest public trust of any item in Q2.

**For the AI industry.** Trust is not absent — it is conditional.
Across all four attitudinal clusters, including the Enthusiasts, UK
Government is named more often than tech companies as the right
governance anchor. Industry credibility flows *through* a regulatory
frame, not around it.

**For consumers and the public conversation.** The interesting public
disagreement is not optimist vs pessimist on the technology itself —
it is whether AI's benefits will be broadly shared. That is the debate
the survey suggests communicators, journalists and policymakers
should be having.

## What surprised us

We came into this analysis expecting one of two pre-committed stories:
either the traditional pattern (DE most worried, ABC1 relaxed) or a
clean inversion (knowledge workers now most worried). The data
supports neither. The honest answer is: **anxiety has become a
consensus, optimism has not.** That is a more nuanced finding than
either pre-committed thesis — and arguably the more important one.

---

# Limitations

- **Frequency ≠ purpose.** Q1 captures how often people use AI, not
  what for. We can describe who uses AI more; we cannot explain why.
  Any functional interpretation (study tool vs emotional support vs
  entertainment) is a hypothesis requiring follow-up research.
- **Cluster stability.** Silhouette is highest at k = 2 (≈ 0.27); we
  retained k = 4 for communication value. Any claim that "there are
  four publics in Britain" overstates the statistical evidence —
  better framed as "four useful personae capturing the attitudinal
  poles and the middle ground."
- **Scale reversal.** The brief mentions a split-sample reversal in
  Q2 that is not visible in the data — we assume it has been applied
  upstream. If that is wrong, Q2 findings need re-running.
- **Weighting.** Design weights exist but are not applied. A final
  submission should apply weights.
- **Theme dictionary.** The Q5 theme tagging is keyword-based and
  transparent, but blunt — a response mentioning "I don't worry
  about jobs" will be tagged as a jobs concern. Hand-validation of a
  subsample would tighten this.
- **Ethnicity sample.** The minority-ethnic sample is ≈85 respondents
  (~15% of n = 569). Enough for headline chi-square tests, too small
  for reliable sub-group cuts. Findings are directional.

---

*End of v2 working draft.*
